Random Generation of Rich \\ Abstract Data Type Values

### 2019-01-29 paper/synthesis.tex
WIP

Characterizing the distribution of values of an arbitrary random generator is a
hard task.
%
It requires to model every random choice that a generator could possibly make to
generate a value.
%
In a recent work \cite{DBLP:conf/haskell/MistaRH18}, we have shown that it is
possible to \emph{analitically} predict the average distribution of data
constructors produced by a random generators automatically derived considering
only ADT definitions---like the one presented on Section
\ref{sec:randomtesting}.
%
For this purpose, we found that random generation of ADT values can be
characterized using the theory of \emph{branching processes} \cite{gw1875}.
%
\footnote{
%
This probabilistic theory was originally conceived to predict the growth and
extinction of royal family trees the Victorian Era, being later applied to a
wide variety of research areas.}
%
In this work we extend this model to predict the average distribution of values
of random generators derived considering structural information coming from
functions pattern matchings and abstract interfaces.

Essentially, a branching process is a special kind of Markov process that models
the evolution of a population of \emph{individuals of different kinds} across
discrete time steps known as \emph{generations}.
%
Each kind of individual is expected to produce an average number of offspring of
(possibly) different kinds from one generation to the next one.
%
In our original setting, we shown that each different data constructor can be
considered as an individual on its own.
%
On the other hand, any ADT value can be seen as a tree where each node
represents a root data constructor and has its sub-expressions as
sub-trees---hence note the similarity with family trees.
%
In this light, each tree level of a random value can be seen as a generation of
individuals in this model.

Then, we can characterize the distribution of constructors that a random
generator produces in the $n$-th generaton as a vector $G_n$ that groups the
number of constructors of each kind produced in that generation.
We note $G_n$ to the vector of real numbers that represents the distribution of
%
Then, the we can predict the expected distribution of constructors at the
generation $n$ (noted as $E[G_n]$) using the following characteristic equation
from branching processes theory:
%
\begin{align}
  E[G_n]^T = E[G_0]^T \cdot M^n
  \label{eqn:prediction}
\end{align}

Where $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given level, provided that we
generated a constructor of kind $i$ at the previous level.
%
Is in this sense that this matrix encodes the ``branching'' behavior of our
random generation from one generation to the next one.
%
In our previous work, we show how each element of this matrix can be calculated
by exploiting type information from data type definitions.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level (corresponding to the element $M_{Join, Text}$) results $2\cdot p_{Text}$,
where $2$ is the number of the recursive fields of type |Html| present on the
constructor |Join| and $p_{Text}$ is the probability of generating the
constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

\todo[inline]{Complete this!}

@@

Our realization in this work is that we can also consider that each different
pattern matching value and abstract interface function call can be considered as
another kind of individual belonging to the random generation process.
%
In this light, we can extend our mean matrix with this information, adding a row
and a column for each different pattern matching value and function call that we
add to the generation process, as we illustrate as follows:

@@

Furthermore, we found that the prediction process can be factored in terms of
two vectors $\beta$ and $\mathcal{P}$, such that $\beta$ represents the number
of recursive sub-terms of each construction that we generate, whereas
$\mathcal{P}$ simply represents the probability of generating that construction
(this probability can be easily calculated based on the generation frequencies
used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}

The vector $\beta$ can be obtained by analyzing the shape of each source of
structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as follows (where dashed lines indicate the composition of the
diffent sub-vectors that can be calculated separately):

>>>>

Characterizing the distribution of values of an arbitrary random generator is a
hard task.
%
It requires to model every random choice that a generator could possibly make to
generate a value.
%
In a recent work \cite{DBLP:conf/haskell/MistaRH18}, we have shown that it is
possible to \emph{analitically} predict the average distribution of data
constructors produced by a random generators automatically derived considering
only ADT definitions---like the one presented on Section
\ref{sec:randomtesting}.
%
For this purpose, we found that random generation of ADT values can be
characterized using the theory of \emph{branching processes} \cite{gw1875}.
%
This probabilistic theory was originally conceived to predict the growth and
extinction of royal family trees the Victorian Era, being later applied to a
wide variety of research areas.
%
In this work, we adapt this model to also predict the average distribution of
values of random generators derived considering structural information coming
from functions pattern matchings and abstract interfaces.

Essentially, a branching process is a special kind of Markov process that models
the evolution of a population of \emph{individuals of different kinds} across
discrete time steps known as \emph{generations}.
%
Each kind of individual is expected to produce an average number of offspring of
(possibly) different kinds from one generation to the next one.
%
Mista el at. \cite{DBLP:conf/haskell/MistaRH18} show that branching processes
can be adapted to predict the generation of ADT values by simply considering
each data constructor as an individual on its own.
%
In fact, any ADT value can be seen as a tree where each node represents a root
data constructor and has its sub-expressions as sub-trees---hence note the
similarity with family trees.
%
In this light, each tree level of a random value can be seen as a generation of
individuals in this model.

We characterize the numbers of constructors that a random generator produces in
the $n$-th generaton as a vector $G_n$, a vector that groups the number of
constructors of each kind produced in that generation---in our |Html| example,
this vector has four components, i.e., one for each constructor.
%
%We note $G_n$ to the vector of real numbers that represents the distribution of
%
From branching processes theory, the following equation captures the expected
distribution of constructors at the generation $n$, noted $E[G_n]$, as follows:
%
\begin{align}
  E[G_n]^T = E[G_0]^T \cdot M^n
  \label{eqn:prediction}
\end{align}

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given level, provided that we
generated a constructor of kind $i$ at the previous level.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions \cite{DBLP:conf/haskell/MistaRH18}---specially the recursive
fields---as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{Join, Text} = 2\cdot p_{Text} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{Text}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

%\todo[inline]{Complete this!}

@@

In this work, we show how to naturally fit structural information beyond ADT
definitions into the prediction mechanism of branching processes.
%
It is our realization that it suffices to simply consider \emph{each different
  pattern matching and functional call as a kind of individual on its own}.
%
In that manner, we can extend our mean matrix $M$ by adding a row and a column
for each different pattern matching and function call as follows:

@@

Furthermore, we found that the prediction process can be factored in terms of
two vectors $\beta$ and $\mathcal{P}$, such that $\beta$ represents the number
of recursive sub-terms of each construction that we generate, whereas
$\mathcal{P}$ simply represents the probability of generating that construction
(this probability can be easily calculated based on the generation frequencies
used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape of each source of
structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as follows (where dashed lines indicate the composition of the
diffent sub-vectors that can be calculated separately):
